Headgroup and glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-06-29 Manuscript/manuscriptPS.tex
Corrections to words and phrases, small cosmetic changes.

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS mediated signaling processes.

@@

Several classical molecular dynamics (MD) simulation studies are done
to understand PS headgroups, their influence on lipid bilayer properties and
interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,venable13,pan14,vangaveti14,melcrova16}.
However, the recent comparisons of PC lipid headgroup and glycerol backbone
C-H bond order parameters calculated from different simulation models
revealed that improvements in the current force fields are needed
to correctly reproduce the headgroup structure and ion binding to
lipid bilayers \cite{botan15,catte16,ollila16}. The ion binding affinity
to POPC bilayer was then improved by implicitly including the electronic
polarizability using the electronic continuum correction \cite{melcr18}.
Here, we collect the set of experimentally measured lipid headgroup and
glycerol backbone C-H bond order parameters, which can be used to
evaluate the quality of headgroup structure and the ion binding affinity
in MD simulations of lipid bilayers containing PS lipids. 
The available MD simulation models of PS are then compared against
the collected experimental data. The results pave the way for the
development of MD simulation force fields that correctly describe
PS lipid headgroup structure and its interactions with ions.
Such models are expected to be useful in elucidating the biological
role of PS and other lipid headgroups because glycerol backbone and
lipid headgroups behave similarly in model membranes and in
bacteria \cite{gally81,scherer87,seelig90}.

@@

   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges)

@@

 lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\

@@

\begin{table*}[!htb]
%\begin{sidewaystable*}[!p]
\centering
\caption{  List of MD simulations. The salt concentrations calculated as 
   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{mukhopadhyay2004})
   and CKP refers to the version with more Gromos compatible version
   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
   %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
   %   The lipid force fields named as in our previous work~\cite{botan15}.
}\label{IONsystems}
    \todo{Berger simulations with NaCl and CaCl to be added}
%begin{minipage}[t]{\textwidth}
\begin{tabular}{l c c r r r r r r c c}
  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
  \hline
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
%    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
%     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}   &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
%    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}    &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    \hline
    POPC:POPS\todoi{This is also probably OPPS? These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
   POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &100 & 120:24 & 5760 & 10  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &300 & 120:24 & 5760 & 31  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &1000 & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    \hline
\end{tabular}
%\end{minipage}
%\end{sidewaystable*}

@@

\subsection{Molecular dynamics simulations}
Molecular dynamics simulation data was collected using
the Open Collaboration method \cite{botan15}.
The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
were used as the communication platforms.
The simulated systems are listed in Table \ref{IONsystems} and
simulation details are given in the SI.
The simulation data is also indexed in the
searchable database (\url{nmrlipids.fi}),
and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).

@@

The order parameters of $\alpha$ and $\beta$ carbons in PC lipids
can be used to measure the ion binding affinity because they 
decrease proportionally to the amound of bound positive
charge to a bilayer \cite{akutsu81,altenbach84,seelig87}.
This molecular electrometer concept is especially useful for 
the comparison between simulations and experiments because
the headgroup order parameters can be directly calculated from 
simulations~\cite{catte16}. Also the headgroup order parameters
of negatively charged PS and PG lipids exhibit systemic, but less
characterized dependence on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers (see supplementary information).

@@

\subsection{Headgroup and glycerol backbone order parameters POPS from $^{13}$C NMR}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
Since only the absolute values of the PS headgroup order parameters were measured
previously \cite{browning80,roux91}, we used the S-DROSS experiment \cite{gross97} to determine the signs of
the order parameters. The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
order parameter is negative, which is confirmed by SIMPSON simulations.
The beginning of the S-DROSS slice suggests that the higher order parameter of
the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter. The S-DROSS curve from
SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.
\begin{figure}[!htb]
  \centering
  \includegraphics[width=8.5cm]{../scratch/figIDEA.eps}
  \caption{\label{PShgSIGNSsimpson}
    (a) The headgroup region of the INEPT spectrum with headgroup and glycerol backbone carbons assigned.
    (b) 2D R-PDLF spectra for headgroup and glycerol backbone regions.
    (c) Slices for $\alpha$ and $\beta$ barbons.
    (d) Experimental SDROSS data (points) and SIMPSON simulations (lines).
    Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the larger and smaller $\alpha$-carbon slittings were used in the
    SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller order parameter (dashed grey).
  }
  \todo{This is preliminary figure, should be polished.}
  \todo{Should we show slices for all the analyzed carbons in (c)?}
\end{figure}

@@

\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters for PS headgroup and glycerol
    backbone from simulations with different models and experiments without CaCl$_2$.
    All DOPS data at 303~K, POPS at 298~K.
    Experimental data from \cite{browning80} contain 0.1~M of NaCl.
    Signs are taken from experiments for POPS described in Supplementary Information.
    The vertical bars shown %for most computational values
    are not error bars, but demonstrate that %for these systems
    we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

@@

None of the tested models reproduce the experimental headgroup and 
glycerol backbone order parameters of DOPS and POPS within the
experimental error bars (Fig. \ref{HGorderParametersPS}). 
The tested models perform generally less well than the models tested
for PC headgroup in the previous study (Fig. 2 in Ref.~\cite{botan15}),
which is also evindent from the comparison between subjective rankings
of the model quality for the headgroup and glycerol backbone 
(Fig. \ref{comparisonTablePS} and Fig. 4 in Ref.~\cite{botan15}).
Therefore, the models cannot be straightforwardly used to interpret the
structural differences between PC and PS headgroups. 
However, the differences are partially reproduced by the two best performing 
models for the $\alpha$ and $\beta$-carbons of PS headgroups, Slipids and CHARMM36.
Both reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order 
in parameter in the PS headgroups (Fig. \ref{HGorderParametersPS} and Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the dihedral angle distributions 
%of CHARMM36 and Slipids in 
in these two models share significant similarities in the headgroup region.  % (Fig.~\ref{dihedralsHG}). 
\todo{Notation of dihedrals in Fig.~\ref{dihedralsHG} should be made somehow combatible with the
chemical structures in Fig.~\ref{HGorderParameters} and the discussion should be then finished.}
However, the glycerol backbone order parameters in Slipids model
significantly differ from CHARMM36 results and experiments (Fig. \ref{HGorderParametersPS}),
which can be related to the differences in the dihedral angle distributions 
of C1-C2-C3-O31 and C2-C3-O31-C31. %(Fig.~\ref{dihedralsGLY}). 
Similar difference was previously observed for PC lipids \cite{botan15}
and its conformational differences  %are illustrated in Figure \ref{glycerol_buslaev}. 
(see supplementary information).

@@

\subsection{Counterion binding to lipid bilayers containing PS lipids}
Membranes containing PS lipids are always accomppanied with counterions, which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models (Fig. \ref{NAdensPOPS}).
The strongest counterion binding, i.e., the lowest concentrations in bulk water,
are observed in MacRog, Berger and Lipid17/JC simulations.
CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
density, while a single maxima is observed in the other models. 
\todo{More detailed discussion may be possible after comparing monovalent ion binding
to bilayers between CHARMM simulations and experiments.}
Area per lipid is in agreement with experiments \cite{pan14} only
in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
%The proposed trend is observed for Lipid17 model,
%for which the Joung-Cheatham ions give higher affinity and smaller area
%per molecule. However, the trend is not observed when comparing accross different
%force fields. For example, CHARMM36ua and MacRog give similar area per molecule
%but binding affinity is significantly higher in MacRog. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

>>>>

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some domains specifically interact with PS lipids,
while others are attracted by general electrostatics 
with a possible regulation by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS mediated signaling processes.

@@

Several classical molecular dynamics (MD) simulation studies are done
to understand PS headgroups, their influence on lipid bilayer properties and
interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,venable13,pan14,vangaveti14,melcrova16}.
However, the recent comparisons of PC lipid headgroup and glycerol backbone
C-H bond order parameters calculated from different simulation models
revealed that improvements in the current force fields are needed
to correctly reproduce the headgroup structure and ion binding to
lipid bilayers \cite{botan15,catte16,ollila16}. The ion binding affinity
to a POPC bilayer was then improved by implicitly including the electronic
polarizability using the electronic continuum correction \cite{melcr18}.
Here, we collect the set of experimentally measured lipid headgroup and
glycerol backbone C-H bond order parameters, which can be used to
evaluate the quality of headgroup structure and the ion binding affinity
in MD simulations of lipid bilayers containing PS lipids. 
The available MD simulation models of PS are then compared against
the collected experimental data. The results pave the way for the
development of MD simulation force fields that correctly describe
PS lipid headgroup structure and its interactions with ions.
Such models are expected to be useful in elucidating the biological
role of PS and other lipid headgroups as glycerol backbone and
lipid headgroups behave similarly in model membranes and in
bacteria \cite{gally81,scherer87,seelig90}.

@@

   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{mukhopadhyay2004})

@@

 lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with the largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\

@@

\caption{  List of MD simulations with additional salt concentrations. 
   The salt concentrations were calculated as 
   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
   %CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
   %(i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
   %'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{mukhopadhyay2004})
   %and CKP refers to the version with more Gromos compatible version
   %(i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
}\label{IONsystems_conc}
    \todo{Berger simulations with NaCl and CaCl to be added}
  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with the largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
%    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
%     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}   &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
%    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}    &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    POPC:POPS\todoi{This is also probably OPPS? These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
   POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &100 & 120:24 & 5760 & 10  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &300 & 120:24 & 5760 & 31  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &1000 & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\

@@

\subsection{Molecular dynamics simulations}
Molecular dynamics simulation data were collected using
the Open Collaboration method \cite{botan15}.
The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
were used as the communication platforms.
The simulated systems are listed in Tables \ref{IONsystems} and \ref{IONsystems_conc}  and
simulation details are given in the SI.
The simulation data are also indexed in the
searchable database (\url{nmrlipids.fi}),
and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).

@@

The order parameters of $\alpha$ and $\beta$ carbons in PC lipids
can be used to measure the ion binding affinity because they 
decrease proportionally to the amount of bound positive
charge to a bilayer \cite{akutsu81,altenbach84,seelig87}.
This molecular electrometer concept is especially useful for 
the comparison between simulations and experiments as
the headgroup order parameters can be directly obtained from 
the both of them~\cite{catte16}. Also the headgroup order parameters
of negatively charged PS and PG lipids exhibit systemic, but less
characterized dependence on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized by measuring the PC headgroup order parameters from 
mixed bilayers (see supplementary information).

@@

\subsection{Headgroup and glycerol backbone order parameters of POPS from $^{13}$C NMR}

The INEPT and 2D R-PDLF experiments from POPS samples give well resolved spectra for all the
carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
Since only the absolute values of the PS headgroup order parameters were measured
previously \cite{browning80,roux91}, we used the S-DROSS experiment \cite{gross97} to determine the signs of
the order parameters. The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
order parameter is negative, which is confirmed by SIMPSON simulations.
The beginning of the S-DROSS slice suggests that the higher order parameter of
the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter. The S-DROSS curve from
SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.
\begin{figure}[!htb]
  \centering
  \includegraphics[width=8.5cm]{../scratch/figIDEA.eps}
  \caption{\label{PShgSIGNSsimpson}
    (a) The headgroup region of the INEPT spectrum with headgroup and glycerol backbone carbons assigned.
    (b) 2D R-PDLF spectra for headgroup and glycerol backbone regions.
    (c) Slices for $\alpha$ and $\beta$ barbons.
    (d) Experimental SDROSS data (points) and SIMPSON simulations (lines).
    Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the larger and smaller $\alpha$-carbon splittings were used in the
    SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller order parameter (dashed grey).
  }
  \todo{This is preliminary figure, should be polished.}
  \todo{Should we show slices for all the analyzed carbons in (c)?}
\end{figure}

@@

\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters for PS headgroup and glycerol
    backbone from simulations with different models and experiments without CaCl$_2$.
    All DOPS data at 303~K, POPS at 298~K.
    Experimental data from \cite{browning80} contain 0.1~M of NaCl.
    Signs are taken from experiments for POPS described in Supplementary Information.
    The vertical bars shown %for most computational values
    are not error bars, but demonstrate that %for these systems
    we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dots mark their time-weighted average.
  }
\end{figure}

@@

None of the tested models reproduce the experimental headgroup and 
glycerol backbone order parameters of DOPS and POPS within the
experimental error bars (Fig. \ref{HGorderParametersPS}). 
The tested models perform generally less well than the models tested
for PC headgroup in the previous study (Fig. 2 in Ref.~\cite{botan15}),
which is also evindent from the comparison between subjective rankings
of the model quality for the headgroup and glycerol backbone 
(Fig. \ref{comparisonTablePS} and Fig. 4 in Ref.~\cite{botan15}).
Therefore, the models cannot be straightforwardly used to interpret the
structural differences between PC and PS headgroups. 
However, the differences are partially reproduced by the two best performing 
models for the $\alpha$ and $\beta$-carbons of PS headgroups, Slipids and CHARMM36,
which both reproduce the larger forking of the $\alpha$-carbon, 
with Slipids reproducing also the lower order parameter of the $\beta$-carbon 
in the PS headgroups (Fig.~\ref{HGorderParametersPS} and Fig.~2 in Ref.~\citenum{botan15}).
Interestingly, the dihedral angle distributions 
%of CHARMM36 and Slipids in 
in these two models share significant similarities in the headgroup region.  % (Fig.~\ref{dihedralsHG}). 
\todo{Notation of dihedrals in Fig.~\ref{dihedralsHG} should be made somehow combatible with the
chemical structures in Fig.~\ref{HGorderParameters} and the discussion should be then finished.}
However, the glycerol backbone order parameters in Slipids model
significantly differ from CHARMM36 results and experiments (Fig. \ref{HGorderParametersPS}),
which can be related to the differences in the dihedral angle distributions 
of C1-C2-C3-O31 and C2-C3-O31-C31. %(Fig.~\ref{dihedralsGLY}). 
Similar difference was previously observed for PC lipids \cite{botan15}
and its conformational differences  %are illustrated in Figure \ref{glycerol_buslaev}. 
(see supplementary information).

@@

\subsection{Counterion binding to lipid bilayers containing PS lipids}
Membranes containing PS lipids are always accomppanied with counterions, which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested to screen the repulsion between charged lipid headgroups 
in MD simulations and, hence, to reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models (Fig. \ref{NAdensPOPS}).
The strongest counterion binding, i.e., the lowest concentrations in bulk water,
are observed in MacRog, Berger and Lipid17/JC simulations.
CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
density, while a single maxima is observed in the other models. 
\todo{More detailed discussion may be possible after comparing monovalent ion binding
to bilayers between CHARMM simulations and experiments.}
Area per lipid is in agreement with experiments \cite{pan14} only
in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained solely by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
%The proposed trend is observed for Lipid17 model,
%for which the Joung-Cheatham ions give higher affinity and smaller area
%per molecule. However, the trend is not observed when comparing accross different
%force fields. For example, CHARMM36ua and MacRog give similar area per molecule
%but binding affinity is significantly higher in MacRog. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameter $\beta$, 
which decreases with the addition of lithium 
but increases with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}), 
while changes of the order parameter $\alpha$ are in the opposite direction. 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 to be written when we have the density profiles.}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data are from Ref.~\citenum{roux90}.
    The values from counterion-only systems are set as a zero point of the y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{CHARMM36 results for this plot would be highly useful.}
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Density distributions of counterions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
  \todo{Lipid 17 is to be added.}
\end{figure}
